Reparametrization of COM-Poisson Regression Models with Applications in the Analysis of Experimental Data

### 2017-10-11 article.Rnw
Corrections and suggestions indicated by Walmes

  In the analysis of count data seldom the equidispersion assumption is
  suitable, hence the Poisson regression model is inappropriate in many
  situations. The COM-Poisson is a member of the exponential family, has
  the Poisson and geometric distributions as special cases and the
  Bernoulli distribution as a limiting case. It can deal with both under
  and overdispersed count data. Therefore it is a suitable distribution
  to analysis of dispersed count data. In spite of the nice properties
  of COM-Poisson, the location parameter of the distribution does not
  represent the average making difficult interpretation in a regression
  structure. In this paper we present a very simple reparametrization
  based on mean approximation. The estimation and inference is based on
  likelihood and simulation study showed that the maximum likelihood
  estimators are unbiased and consistent for both regression and
  precision parameters. Furthermore, we observe empirical correlation
  between regression and precision parameters close to zero, suggest the
  orthogonality feature in the proposed reparametrization model. We
  presents applications of the proposed model by three case studies with
  over, under and equidispersed counts. Results show improvements over
  the conventional approaches. Therefore, its use in the analysis of
  count data is encouraged. The computational routines for fitting of
  original and reparametrized COM-Poisson models are available in the
  supplementary material.

@@

Count data are variables that assume non-negative integer values and
representing the number of times an event occurs in the observation
unit. This kind of data are common in experimental research such as
number of grains produced by a plant, number of fruits produced by a
tree, number of insects on a particular cell, among others. Since the
seminal paper by \citet{Nelder1972} introduced the class of generalized
linear models (GLM), analysis of count data often consider the Poisson
regression model. These models provide a suitable distribution for a
counting random variable and an efficient Newton scoring algorithm are
used to fit them. Furthermore, the theoretical background for GLMs is
well established in the class of dispersion models \citep{Jorgensen1997}
as a generalization of the exponential family of distributions.

In spite of the advantages of the GLM framework, the Poisson
distribution has only one parameter, which represents the mean, and
also, the variance. This relationship is referred as equidispersion.
However, this assumption can be unsuitable to many situations and when
the Poisson model is applied in these cases the standard errors for the
parameters estimates are inconsistent \citep{Winkelmann1995}.

In practice, overdispersion, when the variance greater than the mean, is
largely reported in the literature and may occur due to the absence of
relevant covariates, heterogeneity of sampling units, different domain
not considered and excess of zero \citep{Hinde1998}. Lesser reported,
but that has received attention from the statistical community, is a
underdispersion case, when the variance is less than the mean. The
process that reduce the variability are not as well known that lead to
extra variability. For the same reason, there a few approaches to deal
underdispersed cases. The explanatory mechanisms to underdispersion may
be related to Poisson process. When the time between events is not
exponential distributed, the number of events can be over or
underdispersed, this motivated the class of duration dependence models
\citep{Winkelmann1995}. Other possibility to explain underdispersion are
order statistics. Suppose $Y_i$ is Poisson distributed then maximum of
$Y_i$ is underdispersed.

The strategies for constructing alternative count distributions are
related with the causes of the non-equidispersion. Particular to
overdispersion case Poisson mixed models are widely applied. One popular
example of this approach is the negative-binomial model, where the
natural is assumed to follow a gamma distribution. However, other
distributions for the random effects may be considered, for example the
model Poisson-Tweedie \citep{Bonat2017} and
Poisson-Inverse-Gaussian. For underdispersion, the Gamma-Count
distribution assumes a gamma distribution for the time between events in
Poisson process, thus it can deal with dispersed counts
\citep{Zeviani2014}. The COM-Poisson distribution is given by a
generalization the Poisson distribution allowed for a non-linear
decrease in ratios of successive probabilities \citep{Shmueli2005}.

The COM-Poisson is a member of the exponential family, has the Poisson
and geometric distributions as special cases and the Bernoulli
distribution as a limiting case. It can deal with both under and
overdispersed count data. Some recently applications of the COM-Poisson
distribution are related by \citet{Lord2010} for analysis of traffic
crash data, \citet{Sellers2010} to model airfreight breakage and book
purchases and \citet{Huang2017} with analysis of attendance data,
takeover birds and cotton bolls count. The main disadvantage of
COM-Poisson distribution is that the location parameter does not
represent the average making difficult interpretation in a regression
structure. \citep{Huang2017} proposed a mean-parametrization COM-Poisson
distribution to avoid this issue, in this approach the mean parameter is
obtained by solving the root of an equation defined as a infinite sum.

The main goal of this article is to proposed a novel COM-Poisson
parametrization based on mean approximation given by
\citet{Shmueli2005}. In this parametrization the probability mass
function is wrote with $\mu$ and $\phi$ as function of original
parameters, where $\mu$ is the mean and $\phi$ is the precision or
dispersion parameter. Opposite the original parametrization, the
proposed parametrization leads to interpretability of the parameters in
a regression model and it is more simple than mean-parametrization gives
by \cite{Huang2017} because in this case it is not necessary find $\mu$
as the root of an equation involving an infinite sum. In simulation
study we show the properties about maximum likelihood estimators in
proposed and

This paper is organized as follows. In \autoref{background} we present
the COM-Poisson distribution and \cite{Huang2017} parametrization. The
proposed reparametrization and assessment of moments approximations is
considered in the \autoref{reparametrization}. In the
\autoref{estimation-and-inference} we show the estimation and inference
methods based on likelihood function whose properties are assess in
\autoref{simulation-study} by simulations. Applications of
reparametrized COM-Poisson regression model are illustrated in
\autoref{case-studies} with three case studies. The analysis are
performed using the software R \citep{Rcore2017}. The source code and
data sets are available in supplementary material\footnote{Available on
  \texttt{\url{http://www.leg.ufpr.br/~eduardojr/papercompanions}}
  \label{papercompanion}.}.

@@

The COM-Poisson distribution generalizes the Poisson distribution in
terms of the ratio between the probabilities of two consecutive values,
with addition of a precision parameter \cite{Sellers2010}. Let $Y$ a
COM-Poisson random variable, then
$$\frac{\Pr(Y=y-1)}{\Pr(Y=y)} = \frac{y^\nu}{\lambda}$$ while for the
Poisson distribution this ratio is $\frac{y}{\lambda}$. This allows the
distribution for overdispersion or underdispersion. The probability mass
function for the COM-Poisson random variables takes the form
\begin{equation}
  \label{eqn:pmf-cmp}
  \Pr(Y=y \mid \lambda, \nu) = \frac{\lambda^y}{(y!)^\nu Z(\lambda, \nu)},
  \qquad y = 0, 1, 2, \ldots
\end{equation},
where $\lambda > 0$, $\nu \geq 0$ and $Z(\lambda, \nu) =
\sum_{j=0}^\infty \frac{\lambda^j}{(j!)^\nu}$ is a normalizing
constant.

The $Z(\lambda, \nu)$ series diverges theoretically only when $\nu=0$
and $\lambda \geq 1$, but numerically for small values of $\nu$ combined
with large of $\lambda$, the sum is so huge it causes
\textit{overflow}. \autoref{tab:convergenceZ} shows the calculated
normalizing constants using a thousand increments, that is,
$\sum_{j=0}^{1000}\lambda^j/(j!)^\nu$ for different of $\lambda$ and
$\phi$ \footnote{For $\lambda=1$ e $\nu=0$ we show Inf, although the
  numerically value is $1000$, since the series clearly diverges.}.

@@

# Remove excessive scientific notation
xt2 <- gsub("e\\+00", "    ", format(xt))
xt2 <- gsub("\\|([0-9].{2}) \\|", "\\|**\\1** \\|", xt2)
print(xtable(xt2, digits = -2,
             caption = caption,
             align = "C|CCCCCC",
             # align = "ccccccc",
             label = "tab:convergenceZ"),
      include.colnames = FALSE,
      tabular.environment = "tabularx",
      width = "\\textwidth",
      add.to.row = list(
          pos = list(0, 0),
          command = c(
              " \\multicolumn{7}{c}{$\\bm{\\lambda}$} \\\\\n",
              "$\\bm{\\nu}$ & 0,5 & 1 & 5 & 10 & 30 & 50 \\\\\n")
      ))

@@

An undesirable feature of COM-Poisson distribution is that the moments
cannot be solved in closed form. \citet{Shmueli2005} and
\citet{Sellers2010} using an asymptotic approximation for
$Z(\lambda,\nu)$, shows that an approximate forms can be closed by
\begin{equation}
  \label{eqn:mean-approx}
  E(Y) \approx \lambda^{1/\nu} - \frac{\nu - 1}{2\nu} \qquad
  \textrm{and} \qquad
  \textrm{var}(Y) \approx \frac{\lambda^{1/\nu}}{\nu}\,,
\end{equation} which is particularly accurate for  $\nu \leq 1$ or
$\lambda > 10$. The authors also shows that the mean-variance
relationship can be approximate by $\frac{1}{\nu}E(Y)$. In
\autoref{reparametrization} we assess the approximations.

A regression formulation for COM-Poisson models was proposed by
\citet{Sellers2010}, using the original parametrization. In this case,
the COM-Poisson regression is $\eta(E(Y_i)) = \log(\lambda_i)$ and the
relationship between $E(Y_i)$ and $\bm{x}_i$ is modelled
indirectly. \citet{Huang2017} shows how the means of COM-Poisson
distributions can be modelled directly. in the \autoref{eqn:pmf-cmp},
\Citeauthor{Huang2017} proposes that the parameter $\lambda$ is a
function of $\mu$ and $\nu$, given by the solution to
\begin{equation*}
  \sum_{j=0}^{\infty} (j - \mu) \frac{\lambda^j}{(y!)^\nu} = 0\,.
\end{equation*}
Thus the mean-parametrized COM-Poisson regression is $\eta(E(Y_i)) =
\log(\mu_i)$. In this article, we presented a alternative
mean-parametrization of COM-Poisson models to avoid the limitations of
original parametrization and complexity of \Citeauthor{Huang2017}
parametrization.

@@

The proposed reparametrization of COM-Poisson models is based on
mean approximation (\autoref{eqn:mean-approx}). We introduced new
parameter $\mu$, using the mean approximation
\begin{equation}
  \label{eqn:repar-cmp}
  \mu = h(\lambda, \nu) = \lambda^{1/\nu} - \frac{\nu - 1}{2\nu}
  \quad \Rightarrow \quad
  \lambda = h^{-1}(\mu, \nu) = \left (\mu +
    \frac{(\nu - 1)}{2\nu} \right )^\nu.
\end{equation}
Precision parameter is taken on the log scale to avoid restrictions on the
parameter space, $\phi = \log(\nu), \phi \in \mathbb{R}$. The
interpretation of $\phi$ are the same to $\nu$, but in other scale. For
$\phi<0$ we have overdispersed case and $\phi>0$ overdispersed, relatives
to Poisson distribution. $\phi=0$ represents the special case that the
COM-Poisson comes down to Poisson.

@@

#-------------------------------------------
# Mean and variance relationship
aux <- expand.grid(
    mu = seq(2, 30, length.out = 50),
    phi = seq(log(0.3), log(2.5), length.out = 50))

@@

In order to assess the accuracy of the moments approximations (
\autoref{eqn:mean-approx}) we presented the quadratic errors of moments
approximations in \autoref{fig:approx-plot}(a) and
\ref{fig:approx-plot}(b). Quadratic errors are obtained by $[\mu -
E(Y)]^2$ for expectation and by $\left[ \frac{\mu}{\exp(\phi)} V(Y)
  \right ]^2$, where
$E(Y)$ and $V(Y)$ are numerically calculated using the moments
definition and dotted lines represents the regions $\nu \leq 1$ and
$\lambda > 10^\nu$ in the $(\mu\,,\phi)$ scale.

<<approx-plot, fig.height=3, fig.width=8, fig.cap="Quadratic errors for expectation (a) and variance (b) approximation, dotted lines representing the restriction for good approximations by \\cite{Shmueli2005}. Mean-variance relationship (c), dotted line represents $E(Y)=V(Y)$.">>=

@@

#-------------------------------------------
# Mean-variance relantionship
xy3 <- xyplot(var ~ mean,
              groups = phi,
              data = grid,
              type = "l",
              lwd = 2,
              axis = axis.grid,
              xlab = expression(E(X)),
              ylab = expression(V(X)),
              sub = "(c)",
              legend = list(
                  top = list(
                      fun = draw.colorkey,
                      args = list(
                          key = list(
                              space = "top",
                              col = myreg(length(unique(grid$phi))),
                              at = unique(grid$phi),
                              draw = FALSE)))),
              par.settings = modifyList(
                  ps2, list(superpose.line = list(
                                col = myreg(length(unique(grid$phi))))
                            )
              ),
              panel = function(x, y, ...) {
                  panel.xyplot(x, y, ...)
                  panel.curve(1*x, min(x), max(x), lty = 2)
              },
              page = function(...) {
                  grid.text(expression(phi),
                            just = "bottom",
                            x = unit(0.15, "npc"),
                            y = unit(0.83, "npc"))
              })

@@

The results in \autoref{fig:approx-plot} shows that the mean
approximation is accurate, see maximum quadratic errors is
\Sexpr{max(grid[["emu"]])}. For the variance approximation the maximum
is \Sexpr{max(grid[["eva"]])} and this occurs to negative values of
$\phi$.  Interestingly, the errors are bigger to negative values of
$\phi$ and appear to have no relation with $\mu$ against the regions
gives by \citet{Shmueli2005} ($\phi \leq 0$ and
$\mu > 10 - \frac{\exp(\phi) - 1}{2\exp(\phi)}$). The relationship
between mean and variance is linear with slope controlled by precision
parameter $\phi$ (\autoref{fig:approx-plot}(c)).

The results presented in \autoref{fig:approx-plot}(a) support the
proposed reparametrization. Replacing $\lambda$ and $\nu$ as function of
$\mu$ and $\phi$ in \autoref{eqn:pmf-cmp} the reparametrized
distribution takes the form
\begin{equation}
  \label{eqn:pmf-cmpmu}
  \Pr(Y=y \mid \mu, \phi) =
  \left ( \mu +\frac{ e^\phi-1}{2e^\phi} \right )^{ye^\phi}
  \frac{(y!)^{-e^\phi}}{Z(\mu, \phi)},
  \qquad y = 0, 1, 2, \ldots\,
\end{equation} where $\mu > 0$. We denotes this distribution as
COM-Poisson$_\mu$. In \autoref{fig:pmf-cmp} we show the shapes of
COM-Poisson$_\mu$ distribution.

@@

In this section we shall describe the estimation and inference methods
adopted in this paper. The two forms of COM-Poisson models are fitted by
maximization of the log-likelihood function. The log-likelihood function
based on a set of independent observation $y_i$, $i=1,2,\ldots,n$ for
COM-Poisson$_\mu$ has the following form,
\begin{equation}
  \label{eqn:ll-rcmp}
  \ell = \ell(\bm{\beta}, \phi \mid \bm{y}) =
  e^\phi \left [
    \sum_{i=1}^n y_i
    \log \left( \mu_i + \frac{e^\phi-1}{2e^\phi} \right ) -
    \sum_{i=1}^n \log(y_i!) \right ] -
  \sum_{i=1}^n \log(Z_i(\mu_i, \phi)),
\end{equation}
where $\mu_i = \exp(\bm{x}_i^\top\bm{\beta})$, with
$\bm{x}_i^\top = (x_{i1},\, x_{i2},\, \ldots,\, x_{ip})$ the vector of
known covariates for the $i$-th observation, and $(\bm{\beta},\, \phi)
\in \mathbb{R}^{p+1}$. The normalizing constant $Z_i(\mu_i, \phi)$ is
given by
\begin{equation*}
  Z_i(\mu_i, \phi) = \sum_{j=0}^\infty \left [ \left (
    \mu_i + \frac{e^\phi - 1}{2e^\phi} \right )^{je^\phi}
  \frac{1}{(j!)^{e^\phi}} \right ]
\end{equation*}
To compute the log-likelihood function we need the calculation of an
infinite series for each observation, which makes the computation
expensive for parametric space regions that the convergence of sum is
slow.

@@

The parameter estimation for COM-Poisson model in the original
parametrization is analogous to presented for COM-Poisson$_\mu$, however
considering \autoref{eqn:ll-rcmp} in terms of $\lambda$. Even for the
standard COM-Poisson distribution, the precision parameter is taken on
the log scale, to avoid restriction of the parametric space.

@@

In this section we conducted a simulation study to assess the properties
of the proposed model and estimators. We considered average counts
varying from 3 to 27 according to a regression model structure with a
continuous variable $\bm{x}_1$ between 0 and 1 and a categorical
variable $\bm{x}_2$ with three levels. Then the systematic component is
given by $\bm{\mu} = \exp(\beta_0 + \beta_1 \bm{x}_1 + \beta_{21}
\bm{x}_{21} + \beta_{22} \bm{x}_{22})$, where $\bm{x}_{21}$ and
$\bm{x}_{22}$ are \textit{dummy} variable for $\bm{x}_2$. With respect
to the precision parameter $\phi$, we fixed it to four values $-1.6$,
$-1.0$, $0.0$ and $1.8$, in order to have strong overdispersed, moderate
overdispersed, equidispersed and underdispersed counts
respectively. \autoref{fig:justpars} shows the variation of the average
counts (left) and variation of dispersion index (E$(Y)$ / var$(Y)$) for
each fixed value of the precision parameter $\phi$. This configuration
allows us to assess the properties of the estimators in extreme
situation such as This configuration allows us to evaluate the
properties of the estimators in extreme situations such as high counts
and low dispersion, and low counts and high dispersion.

In order to assess the consistency of the estimators we considered four
different sample sizes, 50, 100, 300 and 1000, generating 1000 datasets
in each case. In \autoref{fig:bias-plot} we show the bias of the
estimators for each scenario (combination between values of the
precision parameter and samples sizes) together with a confidence
interval calculated as average bias plus and minus $\Phi(0.975)$ times
the standard error. The scales are standardized for each parameter by
dividing the bias by the standard error obtained for the sample of size
50.

@@

<<desc-soy, fig.pos="H", fig.height=4, fig.width=8, fig.cap="Number of grains per pot for each potassium dose and humidity level (left) and sample mean agains sample variance of the five replicates for each experimental treatment (right).">>=

>>>>

  In the analysis of count data seldom the equidispersion assumption is
  suitable, hence the Poisson regression model is inappropriate in many
  situations. The COM-Poisson is a member of the exponential family, has
  the Poisson and geometric distributions as special cases and the
  Bernoulli distribution as a limiting case. It can deal with both under
  and overdispersed count data. Therefore it is a suitable distribution
  to the analysis of dispersed count data. In spite of the nice
  properties of COM-Poisson, the location parameter of the distribution
  does not represent the conditional mean making difficult
  interpretation in a regression structure. In this paper we present a
  very simple reparametrization based on mean approximation. The
  estimation and inference is based on likelihood and simulation study
  showed that the maximum likelihood estimators are unbiased and
  consistent for both regression and precision parameters. Furthermore,
  we observed empirical correlation between regression and precision
  parameters close to zero suggesting the orthogonality feature in the
  proposed mean reparametrized model. We present applications of the
  proposed model through three case studies with over, under and
  equidispersed counts. Results show improvements over the conventional
  approaches. Therefore, its use in the analysis of count data is
  encouraged. The computational routines for fitting of original and
  reparametrized COM-Poisson models are available in the supplementary
  material.

@@

Count data are variables that assume non-negative integer values and
represent the number of times an event occurs in the observation
unit. This kind of data is common in experimental research such as
the number of grains produced by a plant, number of fruits produced by a
tree, number of insects captured by a trap, among others. Since the
seminal paper by \citet{Nelder1972} introduced the class of generalized
linear models (GLM), analysis of count data often consider the Poisson
regression model. This model provides a suitable distribution for a
counting random variable and an efficient Newton scoring algorithm is
used to fit it. Furthermore, the theoretical background for GLMs is
well established in the class of dispersion models \citep{Jorgensen1997}
as a generalization of the exponential family of distributions.

In spite of the advantages of the GLM framework, the Poisson
distribution has only one parameter, which represents the mean, and
also, the variance of the count random variable. This relationship is
referred as equidispersion.  However, this assumption can be unsuitable
to many situations and when the Poisson model is applied in these cases
the standard errors for the parameters estimates are inconsistent
\citep{Winkelmann1995}.

In practice, overdispersion, when the variance greater than the mean, is
largely reported in the literature and may occur due to the absence of
relevant covariates, heterogeneity of sampling units, different domain
size not considered and excess of zero \citep{Hinde1998}. Lesser
reported, but that has received attention from the statistical
community, is a underdispersion case, when the variance is less than the
mean. The processes that reduce the variability are not as well known as
those lead to extra variability. For the same reason, there are a few
approaches to deal underdispersed cases. The explanatory mechanisms to
underdispersion may be related to Poisson process. When the time between
events is not exponential distributed, the number of events can be over
or underdispersed, this motivated the class of duration dependence
models \citep{Winkelmann1995}. Other possibility to explain
underdispersion are order statistics. Suppose $Y_i$ is Poisson
distributed then maximum of $Y_i$ is underdispersed.

The strategies for constructing alternative count distributions are
related with the causes of the non-equidispersion. Particular to
overdispersion case Poisson mixed models are widely applied. One popular
example of this approach is the negative-binomial model, where the mean
is added a random noise that is assumed to follow a gamma
distribution. However, other distributions for the random effects may be
considered, for example the model Poisson-Tweedie \citep{Bonat2017} and
Poisson-Inverse-Gaussian. For underdispersion, the Gamma-Count
distribution assumes a gamma distribution for the time between events,
thus it can deal with dispersed counts \citep{Zeviani2014}. The
COM-Poisson distribution is given by a generalization the Poisson
distribution allowed for a non-linear decrease in ratios of successive
probabilities \citep{Shmueli2005}.

The COM-Poisson is a member of the exponential family, has the Poisson
and geometric distributions as special cases and the Bernoulli
distribution as a limiting case. It can deal with both under and
overdispersed count data. Some recently applications of the COM-Poisson
distribution are related by \citet{Lord2010} for analysis of traffic
crash data, \citet{Sellers2010} to model airfreight breakage and book
purchases and \citet{Huang2017} with analysis of attendance data,
takeover birds and cotton bolls count. The main disadvantage of
COM-Poisson distribution is that the location parameter does not
represent the regression conditional mean making difficult
interpretation in a regression structure. \citet{Huang2017} proposed a
mean-parametrization COM-Poisson distribution to avoid this issue, in
this approach the mean parameter is obtained by solving the root of an
equation defined as a infinite sum.

The main goal of this article is to proposed a novel COM-Poisson
parametrization based on mean approximation given by
\citet{Shmueli2005}. In this parametrization the probability mass
function is written with $\mu$ and $\phi$ as function of original
parameters, where $\mu$ is the mean and $\phi$ is the precision or
dispersion parameter. Opposite the original parametrization, the
proposed parametrization leads to interpretability of the parameters in
a regression model and it is more simple than mean-parametrization gives
by \cite{Huang2017} because in this case it is not necessary find $\mu$
as the root of an equation involving an infinite sum. We conduce a
simulation study to performe the properties of the maximum likelihood
estimators for the proposed mean-parametrized COM-Poisson regression
model.

This paper is organized as follows. In \autoref{background} we present
the COM-Poisson distribution and \cite{Huang2017} parametrization. The
proposed reparametrization and assessement of moments approximations is
considered in the \autoref{reparametrization}. In the
\autoref{estimation-and-inference} we show the estimation and inference
methods based on likelihood function whose properties are assess in
\autoref{simulation-study} by simulations. Applications of
reparametrized COM-Poisson regression model are illustrated in
\autoref{case-studies} with three case studies. The analysis are
performed using the software R \citep{Rcore2017}. The source code and
data sets are available in supplementary material\footnote{Available on
  \texttt{\url{http://www.leg.ufpr.br/~eduardojr/papercompanions}}
  \label{papercompanion}.}.

@@

The COM-Poisson distribution generalizes the Poisson distribution in
terms of the ratio between the probabilities of two consecutive values,
with addition of a precision parameter \cite{Sellers2010}. Let $Y$ be a
COM-Poisson random variable, then
$$\frac{\Pr(Y=y-1)}{\Pr(Y=y)} = \frac{y^\nu}{\lambda}$$ while for the
Poisson distribution this ratio is $\frac{y}{\lambda}$. This allows the
distribution deals with overdispersion or underdispersion cases. The probability mass
function for the COM-Poisson random variables takes the form
\begin{equation}
  \label{eqn:pmf-cmp}
  \Pr(Y=y \mid \lambda, \nu) = \frac{\lambda^y}{(y!)^\nu Z(\lambda, \nu)},
  \qquad y = 0, 1, 2, \ldots,
\end{equation}
where $\lambda > 0$, $\nu \geq 0$ and $Z(\lambda, \nu) =
\sum_{j=0}^\infty \frac{\lambda^j}{(j!)^\nu}$ is a normalizing
constant that depends on both parameters.

The $Z(\lambda, \nu)$ series diverges theoretically only when $\nu=0$
and $\lambda \geq 1$, but numerically for small values of $\nu$ combined
with large of $\lambda$, the sum is so huge it causes
\textit{overflow}. \autoref{tab:convergenceZ} shows the calculated
normalizing constants using a thousand increments, that is,
$\sum_{j=0}^{1000}\lambda^j/(j!)^\nu$ for different values of $\lambda$
and $\phi$ \footnote{For $\lambda=1$ e $\nu=0$ we show Inf, although the
  numerically value is $1000$, since the series clearly diverges.}.

@@

# Remove excessive scientific notation
xt2 <- gsub("e\\+00", "    ", format(xt))
xt2 <- gsub("\\|([0-9].{2}) \\|", "\\|**\\1** \\|", xt2)
print(xtable(xt2, digits = -2,
             caption = caption,
             align = "C|CCCCCC",
             # align = "ccccccc",
             label = "tab:convergenceZ"),
      include.colnames = FALSE,
      tabular.environment = "tabularx",
      width = "\\textwidth",
      add.to.row = list(
          pos = list(0, 0),
          command = c(
              " \\multicolumn{7}{c}{$\\bm{\\lambda}$} \\\\\n",
              "$\\bm{\\nu}$ & 0.5 & 1 & 5 & 10 & 30 & 50 \\\\\n")
      ))

@@

An undesirable feature of COM-Poisson distribution is that the moments
cannot be solved in closed form. \citet{Shmueli2005} and
\citet{Sellers2010} using an asymptotic approximation for
$Z(\lambda,\nu)$, showed that an approximate form can be
\begin{equation}
  \label{eqn:mean-approx}
  E(Y) \approx \lambda^{1/\nu} - \frac{\nu - 1}{2\nu} \qquad
  \textrm{and} \qquad
  \textrm{var}(Y) \approx \frac{\lambda^{1/\nu}}{\nu}\,,
\end{equation} which is particularly accurate for  $\nu \leq 1$ or
$\lambda > 10$. The authors also shows that the mean-variance
relationship can be approximate by $\frac{1}{\nu}E(Y)$. In
\autoref{reparametrization} we assess the accucary of these
approximations.

A regression formulation for COM-Poisson models was proposed by
\citet{Sellers2010}, using the original parametrization. In this case,
the COM-Poisson regression is $\eta(E(Y_i)) = \log(\lambda_i)$ and the
relationship between $E(Y_i)$ and $\bm{x}_i$ is modelled
indirectly. \citet{Huang2017} shows how the means of COM-Poisson
distributions can be modelled directly. In the \autoref{eqn:pmf-cmp},
\Citeauthor{Huang2017} proposes that the parameter $\lambda$ as a
function of $\mu$ and $\nu$, given by the solution to
\begin{equation*}
  \sum_{j=0}^{\infty} (j - \mu) \frac{\lambda^j}{(y!)^\nu} = 0\,.
\end{equation*}
Thus the mean-parametrized COM-Poisson regression is $\eta(E(Y_i)) =
\log(\mu_i)$. In this article, we presented an alternative
mean-parametrization of COM-Poisson models to avoid the limitations of
original parametrization and complexity of \Citeauthor{Huang2017}
parametrization.

@@

The proposed reparametrization of COM-Poisson models is based on
mean approximation (\autoref{eqn:mean-approx}). We introduced new
parameter $\mu$, using the mean approximation
\begin{equation}
  \label{eqn:repar-cmp}
  \mu = h(\lambda, \nu) = \lambda^{1/\nu} - \frac{\nu - 1}{2\nu}
  \quad \Rightarrow \quad
  \lambda = h^{-1}(\mu, \nu) = \left (\mu +
    \frac{(\nu - 1)}{2\nu} \right )^\nu.
\end{equation}
The precision parameter is taken on the log scale to avoid restrictions
on the parameter space, $\phi = \log(\nu)$, $\phi \in \mathbb{R}$. The
interpretation of $\phi$ are the same to $\nu$, but in other scale. For
$\phi<0$ we have the overdispersed case and $\phi>0$ underdispersed,
relatives to Poisson distribution. $\phi=0$ represents the special case
that the COM-Poisson recovers to Poisson.

@@

#-------------------------------------------
# Mean and variance relationship
aux <- expand.grid(
    mu = seq(3, 30, length.out = 50),
    phi = seq(-1.5, 1.8, length.out = 50))

@@

In order to assess the accuracy of the moments approximations
(\autoref{eqn:mean-approx}) we presented the quadratic errors of moments
approximations in \autoref{fig:approx-plot}(a) and
\ref{fig:approx-plot}(b). Quadratic errors are obtained by
$[\mu - E(Y)]^2$ for expectation and by
$[ \mu \exp(-\phi) - \text{var}(Y)]^2$, where $E(Y)$ and var$(Y)$ are
numerically calculated using the moments definitio. The dotted lines
represent the border between the regions $\nu \leq 1$ and
$\lambda > 10^\nu$, in the $(\mu\,,\phi)$ scale.

<<approx-plot, fig.height=3, fig.width=8, fig.cap="Quadratic errors for expectation (a) and variance (b) approximation, dotted lines representing the restriction for good approximations by \\cite{Shmueli2005}. Mean-variance relationship (c), dotted line represents $E(Y)=\text{var}(Y)$.">>=

@@

#-------------------------------------------
# Mean-variance relantionship
xy3 <- xyplot(var ~ mean,
              groups = phi,
              data = grid,
              type = "l",
              lwd = 2,
              axis = axis.grid,
              xlab = expression(E(Y)),
              ylab = expression(var(Y)),
              sub = "(c)",
              legend = list(
                  top = list(
                      fun = draw.colorkey,
                      args = list(
                          key = list(
                              space = "top",
                              col = myreg(length(unique(grid$phi))),
                              at = unique(grid$phi),
                              draw = FALSE)))),
              par.settings = modifyList(
                  ps2, list(superpose.line = list(
                                col = myreg(length(unique(grid$phi))))
                            )
              ),
              panel = function(x, y, ...) {
                  panel.xyplot(x, y, ...)
                  panel.curve(1*x, min(x), max(x), lty = 2)
              },
              page = function(...) {
                  grid.text(expression(phi),
                            just = "bottom",
                            x = unit(0.15, "npc"),
                            y = unit(0.83, "npc"))
              })

@@

The results in \autoref{fig:approx-plot} show that the mean
approximation is accurate, the maximum of the quadratic errors is
\Sexpr{max(grid[["emu"]])} in the range evaluated. For the variance
approximation, the maximum is \Sexpr{max(grid[["eva"]])} and this occurs
to negative values of $\phi$. Interestingly, the errors are bigger to
negative values of $\phi$ and appear to have no relation with $\mu$
against the regions gives by \citet{Shmueli2005} ($\phi \leq 0$ and
$\mu > 10 - \frac{\exp(\phi) - 1}{2\exp(\phi)}$).
\autoref{fig:approx-plot}(c) shows that the relationship
between mean and variance is linear with slope controlled by precision
parameter $\phi$, but not in the form $\mu \exp(-\phi)$. In terms of
moments, this leads to a specification indistinguishable from the
quasi-Poisson.

The results presented in \autoref{fig:approx-plot}(a) support the
proposed reparametrization. Replacing $\lambda$ and $\nu$ as function of
$\mu$ and $\phi$ in \autoref{eqn:pmf-cmp}, the reparametrized
distribution takes the form
\begin{equation}
  \label{eqn:pmf-cmpmu}
  \Pr(Y=y \mid \mu, \phi) =
  \left ( \mu +\frac{ e^\phi-1}{2e^\phi} \right )^{ye^\phi}
  \frac{(y!)^{-e^\phi}}{Z(\mu, \phi)},
  \qquad y = 0, 1, 2, \ldots\,,
\end{equation} where $\mu > 0$. We denotes this distribution as
COM-Poisson$_\mu$. In \autoref{fig:pmf-cmp} we show the shapes of
COM-Poisson$_\mu$ distribution.

@@

In this section we shall describe the estimation and inference methods
adopted in this paper. The two forms of COM-Poisson models are fitted by
maximization of the log-likelihood function. The log-likelihood function
based on a set of independent observation $y_i$, $i=1,2,\ldots,n$ for
COM-Poisson$_\mu$ has the following form,
\begin{equation}
  \label{eqn:ll-rcmp}
  \ell = \ell(\bm{\beta}, \phi \mid \bm{y}) =
  e^\phi \left [
    \sum_{i=1}^n y_i
    \log \left( \mu_i + \frac{e^\phi-1}{2e^\phi} \right ) -
    \sum_{i=1}^n \log(y_i!) \right ] -
  \sum_{i=1}^n \log(Z(\mu_i, \phi)),
\end{equation}
where $\mu_i = \exp(\bm{x}_i^\top\bm{\beta})$, with
$\bm{x}_i^\top = (x_{i1},\, x_{i2},\, \ldots,\, x_{ip})$ the vector of
known covariates for the $i$-th observation, and $(\bm{\beta},\, \phi)
\in \mathbb{R}^{p+1}$. The normalizing constant $Z(\mu_i, \phi)$ is
given by
\begin{equation*}
  Z(\mu_i, \phi) = \sum_{j=0}^\infty \left [ \left (
    \mu_i + \frac{e^\phi - 1}{2e^\phi} \right )^{je^\phi}
  \frac{1}{(j!)^{e^\phi}} \right ]
\end{equation*}
To compute the log-likelihood function we need the computation of an
infinite serie for each observation, which makes the computation
expensive for regions in the parameter space where the convergence of
sum $Z$ is slow.

@@

The parameter estimation for COM-Poisson model in the original
parametrization is analogous to presented for COM-Poisson$_\mu$, however
considering \autoref{eqn:ll-rcmp} in terms of $\lambda$. Even for the
standard COM-Poisson distribution, the precision parameter is taken on
the log scale, to avoid restriction of the parameter space.

@@

In this section we conducted a simulation study to assess the properties
of the proposed model and estimators. We considered average counts
varying from 3 to 27 according to a regression model structure with a
continuous variable $\bm{x}_1$ between 0 and 1 and a categorical
variable $\bm{x}_2$ with three levels. Then the systematic component is
given by $\bm{\mu} = \exp(\beta_0 + \beta_1 \bm{x}_1 + \beta_{21}
\bm{x}_{21} + \beta_{22} \bm{x}_{22})$, where $\bm{x}_{21}$ and
$\bm{x}_{22}$ are \textit{dummy} variable for $\bm{x}_2$. With respect
to the precision parameter $\phi$, we fixed it to four values: $-1.6$,
$-1.0$, $0.0$ and $1.8$; in orderto have strong overdispersed, moderate
overdispersed, equidispersed and underdispersed counts
respectively. \autoref{fig:justpars} shows the variation of the average
counts (left) and variation of dispersion index (E$(Y)$/var$(Y)$) for
each fixed value of the precision parameter $\phi$. This configuration
allows us to evaluate the properties of the estimators in extreme
situations such as high counts and low dispersion, and low counts and
high dispersion.

In order to assess the consistency of the estimators we considered four
different sample sizes: 50, 100, 300 and 1000; generating 1000 datasets
in each case. In \autoref{fig:bias-plot} we show the bias of the
estimators for each scenario (combination between values of the
precision parameter and samples sizes) together with a confidence
interval calculated as average bias plus and minus $\Phi(0.975)$ times
the standard error. The scales are standardized for each parameter by
dividing the bias by the standard error obtained for the sample of size
50.

@@

<<desc-soy, fig.pos="H", fig.height=4, fig.width=8, fig.cap="Number of grains per pot for each potassium dose and moisture level (left) and sample mean agains sample variance of the five replicates for each experimental treatment (right).">>=
